---
title: "Vowel Analysis Final Report"
author: 
  - "Student Name"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 3/14/25
format: html
toc: true
editor: visual
theme: spacelab
---

## Vowel Analysis Final Report

### Load packages

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)

# install.packages("phonR")
library(phonR)
```

## Load data

Load your personal data (make sure you update from P106 -\> your P#)

```{r}
# read in data
P106 <- read_csv("data/P106.csv")

# convert variables to factor where needed
convert_to_factor <- function(df, cols) {
  df[cols] <- lapply(df[cols], as.factor)
  return(df)
}

P106 <- convert_to_factor(P106, c("ppt", "word", "ipa", "arpa", "onset", "offset", "environment", "real_word", "sex", "ethnicity", "birthplace", "home_state", "years_lived_hi", "L1", "exposure_langs_yn", "exposure_langs", "speak_other", "item_num", "rep"))

# remove a couple words you won't be needing
P106 <- P106 %>%
  dplyr::filter(!word %in% c("cot", "caught")) # added dplyr to specify which 'filter' to use

```

Class data:

```{r}
# read in data
all_data <- read_csv("data/DS303_combined.csv")

# convert variables to factor where needed
all_data <- convert_to_factor(all_data, c("ppt", "word", "ipa", "arpa", "onset", "offset", "environment", "real_word", "sex", "ethnicity", "birthplace", "home_state", "years_lived_hi", "L1", "exposure_langs_yn", "exposure_langs", "speak_other", "item_num", "rep"))

# remove a couple words you won't be needing
all_data <- all_data %>%
  dplyr::filter(!word %in% c("cot", "caught"))

```

## Explain the Data

(1 point)

In paragraph form:

-   Describe where the data comes from
-   Summarize the contents of the data (how many observations, participants, items, etc.)
-   Mention any pre-processing steps taken. For example, I pre-processed this data by removing words that were obviously mispronounced before even sending it to you. Then, above, you converted certain variables to factor and removed the words "cot" and "caught", which are not relevant to your investigation. Have you done any additional processing?

This data was collected individually from our classmates in this class. Each student filled out a survey that asked for their age, sex, ethnicity, years the speaker has been in college/university, home state, birth place, years lived, first language, whether the speaker had exposure to another language growing up, which language(s) the speaker was exposed to growing up, the age the speaker learned English, and any other languages spoken. Then each classmate was asked to speak a series of words and phrases into a microphone where they're voices were recorded, processed, and analyzed to be used as our data now. Any out of the ordinary stutters, incorrect words, or miss-said words were cut out to minimize error within the data. Above we also removed the words "cot" and "caught" which were not relevant for this project.

## Variables of Interest

(1 point)

For this project, you will explore and analyze the [**class-wide data set**]{.underline}. In paragraph form:

-   Briefly introduce the purpose of this project
-   Identify and explain your variables of interest
-   State research questions or hypotheses about this data

The purpose of this project is to analyze our classes vowel sounds and compare them to other people in the class along with looking at our variables such as hometown, age, sex, ethnicity, and home state. Some variables of interest I have for this project is home state, ethnicity, sex, and age. I hypothesize that classmates who live in the same home state, are the same sex, similar age, and same ethnicity, will have very similar vowel sounds compared to others who differ in these variables. Out of these four variables I hypothesize that home state, sex, and ethnicity, have the most significant effect on the vowel sounds.

For this exercise I will be looking only at home_state and ethnicity. I hypothesize that both of these variables will have a significant influence on the vowel formants of the speaker both f1 and f2.

## EDA and Vowel Plots

(3 points)

-   Generate two vowel plots using `phonR`: one for your own data, and one for class data

-   In a couple sentences, state your observations. Do you see any patterns or differences?

    -   From looking at my own personal data compared to the class data, I found that my results seems to lineup with the majority for the most part when it comes to spacing and format between them.

-   Include at least one visual that supports your hypothesis/justifies your models below, and explain

    -   My first model I ran predicts F1 (Vowel Height) based on home_state and ethnicity

    -   Second model I ran predicts F2 (Vowel Frontness/Backness) based on home_state and ethnicity

    -   From these two models you can see how home_state has more of an effect on f1 and f2 than ethnicity and especially for Nevada (NV), North Dakota (ND), and Pennsylvania (PA) which show significant shifts in f1 and f2.

    -   Another thing I noticed between the two models was that Asian speakers tend to have higher tongue positions, lower f1, but there is no significant difference in f2 when looking at ethnicity.

-   California is the home_state reference and African American is the ethnicity reference

```{r}
poly.area <- with(P106, vowelMeansPolygonArea(f1, f2, ipa, poly.order = c("i", 
    "ɛ", "ɑ", "o", "u", "æ", "ɪ", "ʌ"), group = ppt))
hull.area <- with(P106, convexHullArea(f1, f2, group = ppt))
rbind(poly.area, hull.area)

with(P106, plotVowels(f1, f2, ipa, plot.tokens = TRUE, pch.tokens = ipa, cex.tokens = 1.2, alpha.tokens = 0.2, plot.means = TRUE, pch.means = ipa, cex.means = 2, var.col.by = ipa, ellipse.line = TRUE, pretty = TRUE))


all_data <- read_csv("data/DS303_combined.csv")

# convert variables to factor where needed
all_data <- convert_to_factor(all_data, c("ppt", "word", "ipa", "arpa", "onset", "offset", "environment", "real_word", "sex", "ethnicity", "birthplace", "home_state", "years_lived_hi", "L1", "exposure_langs_yn", "exposure_langs", "speak_other", "item_num", "rep"))

# remove a couple words you won't be needing
all_data <- all_data %>%
  filter(!word %in% c("cot", "caught"))

# plot
with(all_data, plotVowels(f1, f2, ipa, plot.tokens = TRUE, pch.tokens = ipa, cex.tokens = 1.2, alpha.tokens = 0.2, plot.means = TRUE, pch.means = ipa, cex.means = 2, var.col.by = ipa, ellipse.line = TRUE, pretty = TRUE))
```

```{r}
model_f1 <- lm(f1 ~ home_state + ethnicity, data = all_data)
model_f2 <- lm(f2 ~ home_state + ethnicity, data = all_data)

summary(model_f1)
summary(model_f2)
```

## Model Selection and Justification

(3 points)

-   You will build and analyze **two different statistical models** to investigate the relationship between your predictors and outcome variable

    -   For this analysis, I will build and compare two linear regression models to investigate the relationships between home_state, ethnicity, and vowel formants (f1 and f2).

    -   Linear regression is appropriate because f1 and f2 are continuous dependent variables, making it good for analyzing how categorical predictors (home_state and ethnicity) influence vowel formant values.

-   The two models should differ in some way (e.g., one could include an additional or interaction term while the other does not)

-   What statistical models will you use to investigate the relationship between your predictors and outcome variable? (linear vs. logistic regression? mixed effects model?)

-   Why did you select these models?

-   Which variable(s) are included?

```{r}
model_f1 <- lm(f1 ~ home_state + ethnicity, data = all_data)
model_f2 <- lm(f2 ~ home_state + ethnicity, data = all_data)

summary(model_f1)
summary(model_f2)
```

```{r}
model_inter_f1 <- lm(f1 ~ home_state * ethnicity, data = all_data)
model_inter_f2 <- lm(f2 ~ home_state * ethnicity, data = all_data)

summary(model_inter_f1)
summary(model_inter_f2)
```

## Model Comparisons and Best Fit

(3 points)

-   Build and run both models and display their summaries

-   Compare the two models, assess model fit, and determine the better fitting one

```{r}
model_f1 <- lm(f1 ~ home_state + ethnicity, data = all_data)
model_f2 <- lm(f2 ~ home_state + ethnicity, data = all_data)

summary(model_f1)
summary(model_f2)
```

```{r}
model_inter_f1 <- lm(f1 ~ home_state * ethnicity, data = all_data)
model_inter_f2 <- lm(f2 ~ home_state * ethnicity, data = all_data)

summary(model_inter_f1)
summary(model_inter_f2)
```

```{r}
anova(model_f1, model_inter_f1)
anova(model_f2, model_inter_f2)

AIC(model_f1, model_inter_f1, model_f2, model_inter_f2)

step(lm(f1 ~ 1, data = all_data), scope = ~ home_state * ethnicity, direction = "forward")

step(lm(f1 ~ home_state * ethnicity, data = all_data), direction = "backward")

step(lm(f1 ~ home_state * ethnicity, data = all_data), direction = "both")
```

Between the two models, it seems like model 1 is the preferred model between the two. From the anova test there is no significant difference between the two models. The AIC test also confirms that the simpler model, model 1, is the preferred model between the two because the AIC for both models 1 and 2 don't change.

## Interpretation of Results

(3 points)

-   Interpret coefficients and significance
    -   For f1 we see that individuals from North Dakota (60.47) and Pennsylvania (113.05) tend to have higher f1 values compared to individuals from the reference home state, California. While individuals from Washington and Oklahoma tend to have lower f1 values because there coefficients are in the negative.
    -   Asian speakers tend to have significantly lower f1 values than African Americans, and on the other hand, Hawaiian speakers have a higher f1 value.
-   Explain how the predictor variable(s) influence the outcome
    -   I think the predictor variables may have heavily influenced the outcome due to the lack of data and small sample size. There was a lot of NA and I think if done with a larger sample size the results would be a lot different.

## Discussion and Conclusion

(3 points)

-   Summarize key findings
    -   A key finding from this excercise was that home_state and ethnicity do significantly influence vowel formants (f1 and f2) or so they did within our class.
    -   The interaction model I ran did not significantly effect the results
    -   I think different ethnicity do have a strong influence on vowel formants and the way someone talks compares to others. I also believe regional dialects play a large roll in that as well. I think we can see that even here in Hawaii on different islands.
-   Discuss implications
    -   Where you live and what ethnicity you are does have an effect on how you sound.
-   Mention limitations
    -   There wasn't enough people in our class to gather a large enough sample size to really run this experiment to the fullest. I would be really interested to see if we did this with an entire school or college and find the results. I think another thing to add onto is that only looking at vowel formants may not be enough, I think language has a lot to do with influence as well and maybe broadening the scope when investigating this would also be interesting to see.
